There once was a grid at ol' Carkeek

### 2016-04-22 Manuscript/v0.0/carkeek_grid.tex
refocus introduction

Biodiversity surveillance is being revolutionized by DNA-based detection of organisms from environmental samples. ?(specifically speed and scope of ecological studies).
Many researchers are justifiably cautious about the ?(adoption) of this new form of data.
Their apprehension is rooted in the premise that traditional survey approaches are more accurate because the chain of inference between observation and ecological data is usually short: A researcher sees two swans in Lake Hopatcong and infers the lake is occupied by at least 2 swans.
DNA based surveys, on the other hand, consist of a longer chain of inference:
DNA sequences are reported by a sequencing machine, the machine identifies the sequence of products of a polymerase chain reaction (PCR), PCR amplifies pieces of DNA from a purified genomic DNA sample, DNA is purified (extracted) from an environmental sample, environmental samples contain DNA from organisms present, the organisms present are representative of the biological community about which we wish to make inference. ?( reverse order? tie to concrete example (swans of Lake Hopatcong)).
Clearly, this process is more complex than visual surveys, as the relationship between several steps is complex or unknown.
But consider that the processes ?(behind | underlying) other more widely-used ecological survey techniques are similarly complex, such as bird surveys based on song, or visual identification of fungal spores. When alternate survey approaches are impossible or inefficient, we are more willing to accept any available survey data, regardless of the complexity or uncertainty underlying it. (microbiologists have enthusiastically relied on DNA-based surveys for years for this reason, (though yes, they also do not have the problem of disconnect between individual and cell)).

>>>>

Biodiversity surveillance is being revolutionized by DNA-based detection of organisms from environmental samples. ?(specifically speed and scope of ecological studies). % "The surveillance of biodiversity using DNA-based detection from environmental samples"
While this approach has been used for decades to survey micro-organisms (CITE VENTER), it has more recently become used as a tool for surveying macro-organisms (CITE EARLY EDNA).
The technique is founded on the premises that:
1) organisms shed DNA-containing material into the environment, 
2) the DNA contained in this material breaks down over time, 
which leads to the conclusion that:
(1) the concentration of DNA originating from an organism should decrease with distance from its source.
%there should be higher concentrations of DNA close to the organism that shed it

%If DNA never broke down, there would be equal concentrations of DNA from two individuals that passed through the same environment many years apart. Similarly, there would be no relationship between DNA concentration and distance from a source.

For an effectively immobile organism (e.g. fungal hyphae) 
in an effectively static environmental medium (e.g. soil), 
the relationship between DNA concentration and distance appears straightforward (CITE ???).
One might expect the relationship to break down in a dynamic environment: 
An immobile tree may shed cells into the air, which may be carried by the wind for some distance.
This has been a cause for concern for ecologists interested in surveying aquatic organisms, 
because the robustness of eDNA surveys in aquatic environments depends on the distance that DNA travels from a source. 
This distance will be affected by the nature of the environment: 
Material is less likely to be transported (quickly over) great distances in an ice-covered lake 
than along a turbulent shoreline or a fast-moving stream.
One expects more DNA downstream of a fish in a river than upstream, and

In marine environments, this relationship is further complicated by the fact 
that many organisms have dispersive early life history stages (gametes and larvae) 
that can travel great distances but are difficult to track.
These individuals will obscure the expected relationship between DNA concentration and adult abundance because they 
shed genetic material into the environment and 
they are small enough to be captured directly by environmental sampling protocols.
For example, while a traditional survey of a sessile, intertidal bivalve would 
identify the most biomass along the shore, there may be a greater concentration 
of their genetic material at some distance away during a spawning event, or in 
an ocean current carrying many larvae.

Many researchers are justifiably cautious about the ?(adoption) of this new form of data.
Their apprehension is rooted in the premise that traditional survey approaches are more accurate because the chain of inference between observation and ecological data is usually short: A researcher sees two swans in Lake Hopatcong and infers the lake is occupied by at least 2 swans.
DNA based surveys, on the other hand, consist of a longer chain of inference:
DNA sequences are reported by a sequencing machine, the machine identifies the sequence of products of a polymerase chain reaction (PCR), PCR amplifies pieces of DNA from a purified genomic DNA sample, DNA is purified (extracted) from an environmental sample, environmental samples contain DNA from organisms present, the organisms present are representative of the biological community about which we wish to make inference. ?( reverse order? tie to concrete example (swans of Lake Hopatcong)).
Clearly, this process is more complex than visual surveys, as the relationship between several steps is complex or unknown.
But consider that the processes ?(behind | underlying) other more widely-used ecological survey techniques are similarly complex, such as bird surveys based on song, or visual identification of fungal spores. When alternate survey approaches are impossible or inefficient, we are more willing to accept any available survey data, regardless of the complexity or uncertainty underlying it. (microbiologists have enthusiastically relied on DNA-based surveys for years for this reason, (though yes, they also do not have the problem of disconnect between individual and cell)).
